# Light map

In this vignette, we use light data to estimate the position of the Great Reed Warbler (18LCX) at each stationary period by constructing a probability map. This step can be done based on various methods: see @Basson2016 and @Bindoff2018. 

Here we use an approach based on the simple threshold based method, using a calibration of zenith angle. A more thorough introduction to geolocation with light data can be found on <https://geolocationmanual.vogelwarte.ch/> [@Lisovski2020]. 

We first load the data generated in [Pressure map].

```{r}
load("data/1_pressure/18LX_pressure_prob.Rdata")
```

For the calibration of light data, we need additional information. The second calibration is so short (1 day) with 18LX, so we discard it in this example.

```{r}
lon_calib <- 17.05
lat_calib <- 48.9
tm_calib_1 <- c(pam$sta$start[1], pam$sta$end[1])
# tm_calib_2 <- c(pam$sta$start[nrow(pam$sta)], pam$sta$end[nrow(pam$sta)])
```

## Annotate twilights

To find the time of twilight, we can use `find_twilights()`, a function performing the same task than [`TwGeos::FindTwilight()`](https://rdrr.io/github/slisovski/TwGeos/man/findTwilights.html), but using a matrix representation. This approach is faster but less general (e.g., require regularly spaced light data). If it doesn't work with your data, use `TwGeos::FindTwilight()`.

By default, the threshold is the first and last of light day (i.e., `pam$light$obs>0`). The `shift_k` argument is identical to the `offset` in `GeoLight` functions.

```{r}
shift_k <- 0
twl <- find_twilights(pam$light,
  shift_k = shift_k
)
```

We can visualize the twilight with `TwGeos` visualization functions.

```{r}
raw_geolight <- data.frame(
  Date = pam$light$date,
  Light = pam$light$obs
)
lightImage(
  tagdata = raw_geolight,
  offset = shift_k / 60 / 60
)
tsimagePoints(twl$twilight, offset = 0, pch = 16, cex = 1.2)
tsimageDeploymentLines(raw_geolight$Date,
  lon = lon_calib, lat = lat_calib,
  offset = shift_k / 60 / 60, lwd = 3, col = adjustcolor("orange", alpha.f = 0.5)
)
abline(v = tm_calib_1, lty = c(1, 2), col = "firebrick", lwd = 1.5)
```

If you notice any issue with the centering of the night in this figure, specify manually `shift_k`.

The manual editing is easily performed with TRAINSET. In this case, we must label the datapoints we want to delete. Read more about TRAINSET labelling in [Labelling tracks].

We write the twilight data on a `csv` file which can be opened and edited in TRAINSET.

```{r}
write.csv(
  data.frame(
    series = ifelse(twl$rise, "Rise", "Set"),
    timestamp = strftime(twl$twilight, "%Y-%m-%dT00:00:00Z", tz = "UTC"),
    value = (as.numeric(format(twl$twilight, "%H")) * 60 + as.numeric(format(twl$twilight, "%M")) - shift_k / 60 + 60 * 12) %% (60 * 24),
    label = ifelse(is.null(twl$delete), "", ifelse(twl$delete, "Delete", ""))
  ),
  file = "data/2_light/labels/18LX_light.csv",
  row.names = FALSE
)
# browseURL("https://trainset.geocene.com/")
```

When the labelling is finished, export the file and update the `deleted` field in `twl`.

```{r}
csv <- read.csv(paste0("data/2_light/labels/18LX_light-labeled.csv"))
twl$deleted <- !csv$label == ""
```

```{r}
lightImage(tagdata = raw_geolight, offset = 0)
tsimagePoints(twl$twilight,
  offset = 0, pch = 16, cex = 1.2,
  col = ifelse(twl$deleted, "grey20", ifelse(twl$rise, "firebrick", "cornflowerblue"))
)
abline(v = tm_calib_1, lty = c(1, 2), col = "firebrick", lwd = 1.5)
```

## Calibrate zenith angles

Instead of calibrating the twilight errors in terms of duration, we directly model the zenith angle error.

First, we retrieve the twilight during the calibration period.

```{r}
twl_calib <- subset(twl, !deleted & twilight >= tm_calib_1[1] & twilight <= tm_calib_1[2])
```

We then compute the zenith angle (i.e., elevation of the sun) of the twilight time at the calibration site.

```{r}
sun <- solar(twl_calib$twilight)
z <- refracted(zenith(sun, lon_calib, lat_calib))
```

Finally, we fit a kernel distribution for a relatively smooth bandwidth to account for possible bias.

```{r}
fit_z <- density(z, adjust = 1.4, from = 60, to = 120)
hist(z, freq = F)
lines(fit_z, col = "red")
```

The `adjust` parameter allows to manually set how smooth you want the fit to be. Because the zenith angle error model is fitted with data only at the calibration site and that we are using it for all locations of the bird's journey, it is safer to assume a broader/smoother distribution. 

## Compute stationary periods

Before computing the probability map, we group the twilights by stationary period using activity classify in  [Pressure map | Identifying stationary periods](#identifying-stationary-periods).

```{r}
tmp <- which(mapply(function(start, end) {
  start < twl$twilight & twl$twilight < end
}, pam$sta$start, pam$sta$end), arr.ind = TRUE)
twl$sta_id <- 0
twl$sta_id[tmp[, 1]] <- tmp[, 2]
```

## Compute probability map

We first define a grid on which to compute the probabilities. For ease of comparison with the pressure-derived map, we load the grid size and resolution from `pressure_prob` (computed in [Pressure map])

```{r}
g <- as.data.frame(pressure_prob[[1]], xy = TRUE)
g$layer <- NA
```

Selecting only the unlabeled twilights, we compute the probability of observing the zenith angle of each twilight using the calibrated error function for each grid cell.

```{r}
twl_clean <- subset(twl, !deleted)
sun <- solar(twl_clean$twilight)
pgz <- apply(g, 1, function(x) {
  z <- refracted(zenith(sun, x[1], x[2]))
  approx(fit_z$x, fit_z$y, z, yleft = 0, yright = 0)$y
})
```

Aggregating the probability map of each twilight per stationary period requires some assumptions on the independence/correlation of the twilight errors. Read more about this in [Probability aggregation]. Here, we use a log-linear pooling with a weight of $w=0.1$,

```{r}
w <- 0.1
```

We loop through each stationary period and create a raster map with the aggregated probabilities.

```{r}
light_prob <- c()
for (i_s in seq_len(nrow(pam$sta))) {
  id <- twl_clean$sta_id == pam$sta$sta_id[i_s]
  if (sum(id) > 1) {
    g$layer <- exp(colSums(w * log(pgz[id, ]))) # Log-linear equation express in log
  } else if (sum(id) == 1) {
    g$layer <- pgz[id, ]
  } else {
    g$layer <- 1
  }
  gr <- rasterFromXYZ(g)
  crs(gr) <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"
  metadata(gr) <- list(
    sta_id = pam$sta$sta_id[i_s],
    nb_sample = sum(id)
  )
  light_prob[[i_s]] <- gr
}
```

Finally, we can visualize the probability map for each stationary period.

<details>

<summary>See code for figure </summary>

```{r, warning=F}
li_s <- list()
l <- leaflet(width = "100%") %>%
  addProviderTiles(providers$Stamen.TerrainBackground) %>%
  addFullscreenControl()
for (i_r in seq_len(length(light_prob))) {
  i_s <- metadata(light_prob[[i_r]])$sta_id
  info <- pam$sta[pam$sta$sta_id == i_s, ]
  info_str <- paste0(i_s, " | ", info$start, "->", info$end)
  li_s <- append(li_s, info_str)
  l <- l %>% addRasterImage(light_prob[[i_r]], opacity = 0.8, colors = "OrRd", group = info_str)
}
m <- l %>%
  addCircles(lng = lon_calib, lat = lat_calib, color = "black", opacity = 1) %>%
  addLayersControl(
    overlayGroups = li_s,
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  hideGroup(tail(li_s, length(li_s) - 1))
```
</details>

```{r, echo=F}
m
```

## Save

```{r}
save(twl,
  light_prob,
  z,
  fit_z,
  file = "data/2_light/18LX_light_prob.Rdata"
)
```
