# Pressure map

This chapter covers the main steps to determine the position of a bird from pressure data. This code is a direct implementation of the method introduced in Nussbaumer et al. (<a href="https://doi.org/10.1111/2041-210X.14043" target="_blank">2023a</a>).

## Define geographical and temporal parameters

The first step is to define the parameters used to configure your map (and later the graph). These are added to `tag` object.

```{r 02-setmap}
tag <- tag_set_map(tag,
  extent = c(-100, -68, 0, 40), # coordinates of the map to request (W, E, S, N)
  scale = 2, # request on a 1/2=0.5° grid, coarse, but fast
  known = data.frame(
    stap_id = 1,
    known_lat = 37.286812,
    known_lon = -82.304972
  )
)
```

A few indications to select optimal parameters:

- `extent`: the length (`xmin`, `xmax`, `ymin`, `ymax`) or coordinates (W, E, S, N) on which your map is built. A smaller extent results in faster computation, but careful: if the extent is too small and excludes the true position of the bird, the process will still work, leading you to build a wrong trajectory.
- `scale`: the number of pixels per 1° latitude-longitude. (e.g., scale = 10 corresponds to 0.1°\~10km). We recommend starting with a coarse `scale = 1 or 2` and a large `extent` and refining these when you have a better idea of the trajectory. Use the maximal resolution `scale = 10` only for the final run to keep the code fast during development.
- `known`: the position of the bird at equipment and/or retrieval sites (as well as any known position, e.g. observed in the field). These positions can only be provided at the level of a stationary period. Setting this information significantly speeds up the computation by avoiding computing the likelihood map for these stationary periods and narrowing down the size of the graph.

::: {.alert .alert-info role="alert"}
<h4 class="alert-heading mt-2"><strong>What is a GeoPressureR  `map`?</strong></h4>
In GeoPressureR, a `map` object is a container for a spatio-temporal variable, such as the likelihood of pressure. It is therefore defined by its spatial characteristics (`extent` and `scale`), but also with a temporal dimension in terms of stationary periods. The data is stored as a list of matrices, one matrix (or layer) for each stationary period.

See `map_create()` for technical characteristics of a GeoPressureR `map`.
:::

## Compute pressure maps

We are now ready to create the pressure maps! 

To do so, we must match the pressure timeseries of each stationary period with the surface level pressure dataset of <a href"https://doi.org/10.24381/cds.e2161bac" target="_blank">ERA5-Land hourly</a> for all possible pixels of the maps. 

::: {.alert .alert-warning role="alert"}
<h4 class="alert-heading mt-2">
<strong>How does the [GeoPressureAPI](https://github.com/Rafnuss/GeoPressureAPI) work?</strong>
</h4>

To overcome the challenges of handling the large ERA5 dataset, we perform the mismatch computation on the <a href="https://developers.google.com/earth-engine/" target="_blank">Google Earth Engine</a> server which has access to the <a href="https://developers.google.com/earth-engine/datasets/catalog/ECMWF_ERA5_LAND_HOURLY" target="_blank">ERA5 dataset</a> and directly returns the map of mismatch. 

The <a href="https://github.com/Rafnuss/GeoPressureAPI" target="_blank">GeoPressureAPI</a> is a JSON API which serves as an interface to the GEE server.
:::

The function `geopressure_map()` conveniently performs all the necessary steps, but we outline each step below for a comprehensive understanding.

### Compute mismatch maps

```{r 02-geopressure_map_mismatch, cache = TRUE}
tag <- geopressure_map_mismatch(tag,
  max_sample = 100,
  margin = 20,
  quiet = TRUE # hide progress bar in rmarkdown
)
```

- `max_sample` reduces the computational time by limiting the number of data-points used in the match. This usually only impacts long stationary periods where the position is well defined. During labelling, or when accuracy is not critical, it can be convenient to reduce this number between `50` to `100`.
- `margin` can be reduced to `10`-`20` if your bird does not change elevation level during its stationary period.

::: {.alert .alert-warning role="alert"}
<h4 class="alert-heading mt-2"><strong>Taking a long time to compute?</strong></h4>

This function is the most computationally intensive as it needs to: 

1. Pre-process the data (i.e., downscale, clean, etc..) 
2. Generate the requests on the GeoPressureAPI which return the URLs that can be used on the Google Earth Engine (GEE) server to compute the mismatch maps
3. Send the request on the GEE server
4. Wait for the computation on the GEE server and download the map

A progress bar will update you on the completion status, but the timing can be tricky to apprehend because of the computational optimization used (request and computation in parallel) and variability in the GEE server availability.
:::

This function returns the `tag` with two maps:

-   `tag$map_pressure_mse` $\textbf{MSE}$ is a GeoPressureR `map` of the normalized mean square error between the pressure timeseries and ERA5 map.
-   `tag$map_pressure_mask` $\textbf{z}_{thr}$ is a GeoPressureR `map` of the proportion of data-points in the pressure timeseries which correspond to an altitude that falls between the min and max altitude of each grid cell (accounting for the `margin` parameter).

```{r, warning = F}
plot(tag, type = "map_pressure_mse")
```

```{r, warning = F}
# This is an alternative and equivalent way to plot a map in tag
plot(tag$map_pressure_mask)
```

### Compute likelihood maps

We combine and convert these two maps into a single likelihood map using $$f \propto \exp \left(-w(n) \frac{\textbf{MSE}}{\sigma^2} \right) [\textbf{z}_{thr}>thr_{mask}]$$ where $\sigma$ is the standard deviation of pressure and $thr_{mask}$ is the threshold of the mask. 

Because the auto-correlation of the timeseries is not accounted for in this equation, we use a log-linear pooling weight $w(n)=\log(n)/n$, where $n$ is the number of data-points in the time series. See [Probability aggregation](probability-aggregation.html#gaussian-likelihood-function) for more information. 

```{r}
tag <- geopressure_map_likelihood(
  tag,
  sd = 1,
  thr_mask = 0.9,
  log_linear_pooling_weight = \(n) log(n) / n
)
```

::: {.alert .alert-warning role="alert"}
<h4 class="alert-heading mt-2"><strong>Calibrating `sd`</strong></h4>

The standard deviation `sd` ($\sigma$) plays an important role in the spread of the uncertainty of your map. It accounts for (1) error in reanalysis data (may vary spatially), (2) sensor error (very low), and (3) bird vertical movement (likely most significant, 1hPa = 10m e.g. up and down a tree). 

This value should ideally be calibrated, but it is usually safe to start with a value of `1`. We see how to adjust this value in [check #4 of the labelling procedure](labelling tracks#test-4-histogram-of-pressure-error). 

The parameter $thr_{mask}$ has less influence and can usually be left at `0.9`.
:::

The resulting pressure likelihood map can be visualized with:

```{r, warning = F}
plot(tag$map_pressure)
```
