# Static map

This chapter covers important pre-processing steps to ensure that modelling the trajectory with a graph is possible and successful. Firstly, we will aligned the maps of pressure and light and create a `static_prob` variable containing all the information necessary for modeling the trajectory (incl. flight info).

But before moving too quickly, we need to carefully check that pressure, light and flight duration data allow for a coherent trajectory. It is possible (even probable) that some manual editing of the pressure data labelling will be required, especially for short stopovers.

## Combine pressure and light

Start by loading the data computed in [Light map] and [Pressure map].

```{r}
load("data/1_pressure/18LX_pressure_prob.Rdata")
load("data/2_light/18LX_light_prob.Rdata")
```

We first need to retrieve the pressure and light data for the same stationary period.

When running this code for the first time, it is recommended to start by keeping only long stationary periods (e.g. `thr_sta_dur=5\*24` hours), and when all the check below are passing, you can reduce this threshold to zero.

```{r}
thr_sta_dur <- 0 # in hours
sta_pres <- unlist(lapply(pressure_prob, function(x) raster::metadata(x)$sta_id))
sta_light <- unlist(lapply(light_prob, function(x) raster::metadata(x)$sta_id))
sta_thres <- pam$sta$sta_id[difftime(pam$sta$end, pam$sta$start, units = "hours") > thr_sta_dur]
# Get the sta_id present on all three data sources
sta_id_keep <- intersect(intersect(sta_pres, sta_light), sta_thres)
# Filter pressure and light map
pressure_prob <- pressure_prob[sta_pres %in% sta_id_keep]
light_prob <- light_prob[sta_light %in% sta_id_keep]
```

We then need to keep all the flights between consecutive stationary period separate so that we can estimate the wind support correctly.

```{r}
flight <- list()
for (i_f in seq_len(length(sta_id_keep) - 1)) {
  from_sta_id <- sta_id_keep[i_f]
  to_sta_id <- sta_id_keep[i_f + 1]
  flight[[i_f]] <- list(
    start = pam$sta$end[seq(from_sta_id, to_sta_id - 1)],
    end = pam$sta$start[seq(from_sta_id + 1, to_sta_id)],
    sta_id = seq(from_sta_id, to_sta_id - 1)
  )
}
flight[[i_f + 1]] <- list()
```

We compute the static probability with the product of light and pressure probability maps, and add the flight duration in the metadata. `static_prob` is the consolidate variable containing all the information necessary to run the graph functions.

```{r}
static_prob <- mapply(function(light, pressure, flight) {
  # define static prob as the product of light and pressure prob
  static_prob <- light * pressure

  # replace na by zero
  # tmp <- values(static_prob)
  # tmp[is.na(tmp)] <- 0
  # values(static_prob) <- tmp

  metadata(static_prob) <- metadata(pressure)
  metadata(static_prob)$flight <- flight

  return(static_prob)
}, light_prob, pressure_prob, flight)
```

We overwrite the probability of the first and last stationary periods with the known location of the equipment/retrieval sites.

```{r}
lon_calib <- 17.05
lat_calib <- 48.9

lat <- seq(raster::ymax(static_prob[[1]]), raster::ymin(static_prob[[1]]), length.out = nrow(static_prob[[1]]) + 1)
lat <- lat[seq_len(length(lat) - 1)] + diff(lat[1:2]) / 2
lon <- seq(raster::xmin(static_prob[[1]]), raster::xmax(static_prob[[1]]), length.out = ncol(static_prob[[1]]) + 1)
lon <- lon[seq_len(length(lon) - 1)] + diff(lon[1:2]) / 2

lon_calib_id <- which.min(abs(lon_calib - lon))
lat_calib_id <- which.min(abs(lat_calib - lat))

tmp <- as.matrix(static_prob[[1]])
tmp[!is.na(tmp)] <- 0
tmp[lat_calib_id, lon_calib_id] <- 1
values(static_prob[[1]]) <- tmp

tmp <- as.matrix(static_prob[[length(static_prob)]])
tmp[!is.na(tmp)] <- 0
tmp[lat_calib_id, lon_calib_id] <- 1
values(static_prob[[length(static_prob)]]) <- tmp
```

Finally, we can retrieve the pressure and altitude from the most likely position of the combined map of pressure and light.

```{r, cache=TRUE, results='hide', message=FALSE, warning=FALSE}
path <- geopressure_map2path(static_prob)
static_timeserie <- geopressure_ts_path(path, pam$pressure)
```

## Check track with GeoPressureViz

Now that we have combined pressure and light, we need to verify that the data is coherent.

In [Labelling tracks], we already checked that the pressure timeseries measured by the geolocator are consistent with a least one location on the map. However, we didn't check whether these locations are (1) coherent with light data, and (2) within reach of one another, considering flight duration and realistic flight speed.

To carry out these checks, you can use the shiny app `GeoPressureViz` which helps you visualize the overall trajectory of the bird as well as each step-by-step move. 

```{r, eval = F}
geopressureviz(
  pam = pam,
  pressure_prob = pressure_prob,
  light_prob = light_prob
  static_prob = static_prob, 
  pressure_timeserie = static_timeserie 
)
```

Learn more about this app in the dedicated chapter [GeoPressureViz].

With GeoPressureViz, we want to check that the probability map is coherent with the flight distance from the previous/next location and ultimately that a overal trajectory is possible/present in the data. If you notice any inconsistency, go back to the labellisation step (for pressure and light) and check again.

I suggest starting by selecting only long stopovers (24-72 hours depending on your application), as this will help to draw out the general trajectory followed by the bird, and only later adding the shorter stopovers which could raise more confusion (the best match of pressure is often completely off).

Following the same recommendation than in [Labelling tracks], you want the non-outlier pressure of each stationary period to match a single elevation timeserie (bottom panel). You can check on the map that the pressure location seem coherent by selecting only "pressure" in the probability map display. Usually, it's pretty obvious when there is an issue on the map.

So, to correct any issue, go back to trainset and edit (1) labelling of activity to split or combine stationary periods and (2) labelling the pressure timeserie to exclude certain datapoints from the match. The good match of the timeserie is essential and will require several iteration.

Some additional notes: 

- The distance is computed based on an assumed average groundspeed. Birds can fly with a groundspeed up to 120-150 km/h (with wind support) although their usual average is around 40km/h. 
- The default position of the bird in GeoPressureViz is based on the most likely position from the static probability map. This is (usually) not the correct location. 
- Bird tend to flight directly to their main destination. Most detour are artifact, or you really need a long stationary period with a good match of pressure. 
- The most regular issue I faced is a small vertical movement of the bird during a 2-8 days stopover. Shorter stopover are usually easy as the bird doesn't move much but each species is different.
- Wind data is not yet included and might explain some longer flight (sometimes up to 100km/h).

## Final checks

These checks are performed when creating the graph `graph_create()`, but for pedagogical reason, I thought it would be better to introduce them step by step here as they will also help you to get a better sense of the specific movement/trajectory of the bird you are modeling.

### Check 1

A first and easy check is that there be at least one location with a probability greater than 1 for each stationary period.

```{r}
static_prob_n <- lapply(static_prob, function(x) {
  probt <- raster::as.matrix(x)
  probt[is.na(probt)] <- 0
  probt / sum(probt, na.rm = T)
})
tmp <- unlist(lapply(static_prob_n, sum)) == 0
if (any(tmp)) {
  warning(paste0(
    "The `static_prob` provided has a probability map equal to ",
    "zero for the stationary period: ", which(tmp)
  ))
}
```

### Check 2

Secondly, we check that there always be at least one possible transition from one stationary period to the next.

```{r}
for (i_s in seq_len(length(static_prob) - 1)) {
  cur <- as.matrix(static_prob[[i_s]]) > 0
  cur[is.na(cur)] <- F
  nex <- as.matrix(static_prob[[i_s + 1]]) > 0
  nex[is.na(nex)] <- F

  mtf <- metadata(static_prob[[i_s]])
  flight_duration <- as.numeric(sum(difftime(mtf$flight$end, mtf$flight$start, unit = "hours"))) # hours
  resolution <- mean(res(static_prob[[1]])) * 111 # assuming 1°= 111km
  thr_gs <- 150 # Assuming a max groundspeed of 150km/h

  # Check possible position at next stationary period
  possible_next <- (EBImage::distmap(!cur) * resolution / flight_duration) < thr_gs

  if (sum(possible_next & nex) == 0) {
    stop(paste("There are no possible transition from stationary period", i_s, "to", i_s + 1, ". Check part 1 process (light and pressure)", sep = " "))
  }
}
```

## Save

```{r}
save(pam,
  static_prob,
  static_timeserie,
  file = "data/3_static/18LX_static_prob.Rdata"
)
```
