# Wind graph

In this final chapter, we will add wind data to the graph to refine the transition probability based on an assumed airspeed of the bird rather than groundspeed [@Werfeli2020].

First load the static probability map. 

```{r}
load("data/3_static/18LX_static_prob.Rdata")
```

## Download wind data

Wind data is available at high resolution (1hr, 0.25°, 37 pressure level) on [ERA5 hourly data on pressure levels](https://doi.org/10.24381/cds.bd0915c6). And this data is easily accessible through the [`ecmwfr`](https://bluegreen-labs.github.io/ecmwfr) package. 

The first step is to setup-up your CDS access. You will need to create an account on [https://cds.climate.copernicus.eu/](https://cds.climate.copernicus.eu/user/register) to generate your API key and uid number. You can stored them in your `.Rprofile` with the commented code.

```{r, eval=F}
Sys.setenv(cds_key = "Insert_your_CDS_API_KEY_here")
Sys.setenv(cds_user = "Insert_your_CDS_UID_here")
usethis::edit_r_environ()
```

This will allow you to retrieve them whenever you need with

```{r, eval=F}
Sys.getenv("cds_key")
Sys.getenv("cds_user")
```

As the flight are of short duration, we suggest to download a file for each flight, this can be done automatically with `graph_download_wind` which uses [`wf_request_batch()`](https://bluegreen-labs.github.io/ecmwfr/articles/advanced_vignette.html#batch-parallel-requests) to make all requests in parallel. 

```{r, eval=F}
graph_download_wind(pam,
  area = static_prob, # area accept also c(50, -16, 0, 23)
)
```

If you have any issue with the download, you can monitor the requests at <https://cds.climate.copernicus.eu/cdsapp#!/yourrequests>. In case you have a lot of flight to requests (>20), for instance if you want to download the data for multiple tracks at once. It is best to use `transfer=FALSE` which makes the request and return the URL for the download without waiting for the download to be available. You can then transfer the file using the URL. 

```{r, eval=F}
requests <- graph_download_wind(pam,
  area = static_prob,
  transfer = FALSE
)
lapply(requests, function(x) {
  x$transfer()
})
```

## Create graph

We first create the graph identically to in [basic graph](/articles/basic-graph)

```{r, cache=TRUE, message=F, results='hide'}
grl <- graph_create(static_prob, thr_prob_percentile = .99, thr_gs = 150)
```

## Add wind to graph

We can compute the windspeed experienced by the bird if he had flew each possible transition (i.e. edge in the graph). Based on this windspeed and groundspeed, we also compute the airspeed. All of these are stored as complex value with the real part representing the E-W component and the imaginary part corresponding to the N-S.

```{r, cache=TRUE, message=F, results='hide'}
grl <- graph_add_wind(grl,
  pressure = pam$pressure,
  filename = "data/5_wind_graph/18LX/18LX_",
  thr_as = 100
)
```

## Compute the transition probability

Now that the have computed the airspeed required for performing the transition of each edge, we can improve the computation of the probability by modeling the probability of airspeed rather than groundspeed.

We first search the morphological information of the Great Reed Warbler using the [AVONET database](https://doi.org/10.6084/m9.figshare.16586228.v5). You can also overwrite any of these value if you know them. See `flight_bird()` for more details.

```{r}
bird <- flight_bird("Acrocephalus arundinaceus")
```

Using the bird created, we can convert a airspeed into a probability using the power method in the function `flight_prob()`. See [Basic graph] for more information on `flight_prob()`.
As an example, we can plot the probability for an airspeed ranging between 0 and 80 km/h.

```{r}
speed <- seq(0, 80)
prob <- flight_prob(speed,
  method = "power",
  bird = bird,
  low_speed_fix = 10,
  fun_power = function(power) {
    (1 / power)^3
  }
)
plot(speed, prob, type = "l", xlab = "Airspeed [km/h]", ylab = "Probability")
```

We can then compute the probability of each transition of the graph.

```{r}
grl$p <- grl$ps * flight_prob(grl$as, method = "power", bird = bird, low_speed_fix = 10)
```

## Output 1: Shortest path

Same as [Basic graph]

```{r}
g <- graph_from_data_frame(data.frame(
  from = grl$s,
  to = grl$t,
  weight = -log(grl$p)
))
sp <- shortest_paths(g, from = paste(grl$equipment), to = paste(grl$retrieval))

# Convert igraph representation to lat-lon
grl$shortest_path <- graph_path2lonlat(as.numeric(sp$vpath[[1]]$name), grl)
```

We can visualize the shortest path with the windpseed direction (arrow) and magnitude (color) experienced during this particular flight

<details>

<summary>See code for figure </summary>

```{r}
fun_marker_color <- function(norm) {
  if (norm < 20) {
    "darkpurple"
  } else if (norm < 35) {
    "darkblue"
  } else if (norm < 50) {
    "lightblue"
  } else if (norm < 60) {
    "lightgreen"
  } else if (norm < 80) {
    "yellow"
  } else if (norm < 100) {
    "lightred"
  } else {
    "darkred"
  }
}
fun_NSEW <- function(angle) {
  angle <- angle %% (pi * 2)
  angle <- angle * 180 / pi
  if (angle < 45 / 2) {
    "E"
  } else if (angle < 45 * 3 / 2) {
    "NE"
  } else if (angle < 45 * 5 / 2) {
    "N"
  } else if (angle < 45 * 7 / 2) {
    "NW"
  } else if (angle < 45 * 9 / 2) {
    "W"
  } else if (angle < 45 * 11 / 2) {
    "SW"
  } else if (angle < 45 * 13 / 2) {
    "S"
  } else if (angle < 45 * 15 / 2) {
    "SE"
  } else {
    "E"
  }
}

sta_duration <- unlist(lapply(static_prob, function(x) {
  as.numeric(difftime(metadata(x)$temporal_extent[2], metadata(x)$temporal_extent[1], units = "days"))
}))

m <- leaflet(width = "100%") %>%
  addProviderTiles(providers$Stamen.TerrainBackground) %>%
  addFullscreenControl() %>%
  addPolylines(lng = grl$shortest_path$lon, lat = grl$shortest_path$lat, opacity = 1, color = "#808080", weight = 3) %>%
  addCircles(lng = grl$shortest_path$lon, lat = grl$shortest_path$lat, opacity = 1, color = "#000", weight = sta_duration^(0.3) * 10)

for (i_s in seq_len(grl$sz[3] - 1)) {
  if (grl$flight_duration[i_s] > 5) {
    edge <- which(grl$s == grl$shortest_path$id[i_s] & grl$t == grl$shortest_path$id[i_s + 1])

    label <- paste0(
      i_s, ": ", grl$flight[[i_s]]$start, " - ", grl$flight[[i_s]]$end, "<br>",
      "F. dur.: ", round(grl$flight_duration[i_s]), " h <br>",
      "GS: ", round(abs(grl$gs[edge])), " km/h, ", fun_NSEW(Arg(grl$gs[edge])), "<br>",
      "WS: ", round(abs(grl$ws[edge])), " km/h, ", fun_NSEW(Arg(grl$ws[edge])), "<br>",
      "AS: ", round(abs(grl$as[edge])), " km/h, ", fun_NSEW(Arg(grl$as[edge])), "<br>"
    )

    iconArrow <- makeAwesomeIcon(
      icon = "arrow-up",
      library = "fa",
      iconColor = "#FFF",
      iconRotate = (90 - Arg(grl$ws[edge]) / pi * 180) %% 360,
      squareMarker = TRUE,
      markerColor = fun_marker_color(abs(grl$ws[edge]))
    )

    m <- m %>% addAwesomeMarkers(
      lng = (grl$shortest_path$lon[i_s] + grl$shortest_path$lon[i_s + 1]) / 2,
      lat = (grl$shortest_path$lat[i_s] + grl$shortest_path$lat[i_s + 1]) / 2,
      icon = iconArrow, popup = label
    )
  }
}
```
</details>
```{r, echo=F}
m
```

## Output 2: Marginal probability map

Same as [Basic graph]

```{r}
grl_marginal <- graph_marginal(grl)
```

<details>

<summary>See code for figure </summary>

```{r, warning=F}
li_s <- list()
l <- leaflet(width = "100%") %>%
  addProviderTiles(providers$Stamen.TerrainBackground) %>%
  addFullscreenControl()
for (i_r in seq_len(length(grl_marginal))) {
  i_s <- metadata(static_prob[[i_r]])$sta_id
  info <- metadata(static_prob[[i_r]])$temporal_extent
  info_str <- paste0(i_s, " | ", info[1], "->", info[2])
  li_s <- append(li_s, info_str)
  l <- l %>%
    addRasterImage(grl_marginal[[i_r]], colors = "OrRd", opacity = 0.8, group = info_str) %>%
    addCircles(lng = grl$shortest_path$lon[i_s], lat = grl$shortest_path$lat[i_s], opacity = 1, color = "#000", weight = 10, group = info_str)
}
m <- l %>%
  addPolylines(lng = grl$shortest_path$lon, lat = grl$shortest_path$lat, opacity = .5, color = "#808080", weight = 0.5) %>%
  addCircles(lng = grl$shortest_path$lon, lat = grl$shortest_path$lat, opacity = .5, color = "#000", weight = sta_duration^(0.3) * 10) %>%
  addLayersControl(
    overlayGroups = li_s,
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  hideGroup(tail(li_s, length(li_s) - 1))
```
</details>
```{r, echo=F}
m
```


## Output 3: Simulated paths

Same as [Basic graph].

```{r, results='hide', message=FALSE}
nj <- 10 # Number of simulation
path_sim <- graph_simulation(grl, nj = nj)
```

## Run GeoPressureViz

We can also visualize the shortest path and marginal map in [GeoPressureViz]. First, we need to query the pressure timeserie on the shortest path location. 

```{r, cache = T, message=F, results='hide', warning=FALSE}
shortest_path <- as.data.frame(grl$shortest_path)
shortest_path_timeserie <- geopressure_ts_path(shortest_path, pam$pressure)
```

```{r, eval = F}
geopressureviz(
  pam = pam,
  pressure_prob = pressure_prob,
  light_prob = light_prob,
  static_prob = static_prob,
  pressure_timeserie = shortest_path_timeserie,
  static_prob_marginal = grl_marginal
)
```

## Compute flight energy

We can compute the energy [Joules] expense of each flight for the bird

```{r}
edge <- t(graph_path2edge(path_sim$id, grl))

# Convert airpseed from km/hr to m/s and use the bird constructure earlier to compute the mechanical power [W=J/s]
p_mech <- flight_power(abs(grl$as[edge]) * 1000 / 60 / 60, bird = bird)

# Convert the power into energy [J]
energy <- p_mech * rep(head(grl$flight_duration, -1) * 60 * 60, nj)
dim(energy) <- dim(edge)
```

And compare the histogram of ground, wind, airspeed, flight duration and energy for each of the simulation. 

```{r, fig.height=5}
energy_df <- data.frame(
  energy = as.vector(energy),
  as = abs(grl$as[edge]),
  gs = abs(grl$gs[edge]),
  ws = abs(grl$ws[edge]),
  sta_id_s = rep(head(grl$sta_id, -1), nj),
  sta_id_t = rep(tail(grl$sta_id, -1), nj),
  flight_duration = rep(head(grl$flight_duration, -1), nj)
)
energy_df$name <- paste(energy_df$sta_id_s, energy_df$sta_id_t, sep = "-")

plot1 <- ggplot(energy_df, aes(reorder(name, sta_id_s), gs)) +
  theme_bw() +
  geom_boxplot()
plot2 <- ggplot(energy_df, aes(reorder(name, sta_id_s), ws)) +
  theme_bw() +
  geom_boxplot()
plot3 <- ggplot(energy_df, aes(reorder(name, sta_id_s), as)) +
  theme_bw() +
  geom_boxplot()
plot4 <- ggplot(energy_df, aes(reorder(name, sta_id_s), flight_duration)) +
  theme_bw() +
  geom_point()
plot5 <- ggplot(energy_df, aes(reorder(name, sta_id_s), energy)) +
  theme_bw() +
  geom_boxplot()
# grid.arrange(plot1, plot2, plot3, plot4, plot5, nrow=5)

subplot(ggplotly(plot1), ggplotly(plot2), ggplotly(plot3), ggplotly(plot4), ggplotly(plot5), nrows = 5, titleY = TRUE)
```

## Save
```{r}
save(
  path_sim,
  shortest_path,
  static_prob_marginal,
  shortest_path_timeserie,
  file = "data/5_wind_graph/18LX_wind_graph.Rdata"
)
```
