# Trajectory with wind

In this second chapter of the advance tutorial, we will see how to model the trajectory of the Great Reed Warbler using wind data. 

Wind can significantly contribute to a bird movement, explained up to 50% of the displacement! Accounting for wind allows to estimate the airspeed of each transition rather than groundspeed. As such, the movement model can be defined as the probability of a bird airspeed (rather than groundspeed), which is much more constrained and precise. The general approach on how this is done with the graph is presented in details in [section 2.2.4 of Nussbaumer et al. (2023b)](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.14082#mee314082-sec-0008-title).

## Download wind data

Wind data is available at high resolution (1hr, 0.25°, 37 pressure level) on [ERA5 hourly data on pressure levels](https://doi.org/10.24381/cds.bd0915c6). This data is easily accessible through the [`ecmwfr`](https://bluegreen-labs.github.io/ecmwfr) package. 

The first step is to setup-up your CDS access. You will need to create an account on [cds.climate.copernicus.eu](https://cds.climate.copernicus.eu/user/register) to generate your API key and uid number. You can store them as environment variables in your `.Rprofile` using `Sys.setenv()`.

```{r, eval=F}
Sys.setenv(cds_key = "Insert_your_CDS_API_KEY_here")
Sys.setenv(cds_user = "Insert_your_CDS_UID_here")
usethis::edit_r_environ() # opens .Rprofile
```

You can then retrieve them whenever you want with:

```{r, eval=F}
Sys.getenv("cds_key")
Sys.getenv("cds_user")
```

As the flights tend to be of short duration, we suggest to download a file for each flight. This can be done automatically with `graph_download_wind()`, which uses [`wf_request_batch()`](https://bluegreen-labs.github.io/ecmwfr/articles/advanced_vignette.html#batch-parallel-requests) to make all the requests in parallel. 

```{r, eval=F}
tag_download_wind(tag)
```

You can monitor the requests at <https://cds.climate.copernicus.eu/cdsapp#!/yourrequests>.

In case you have a lot of tracks for which you need to download wind data and don't want to block your console, you might consider using an [RStudio background job](https://solutions.rstudio.com/r/jobs/), which can be easily called with the [job package](https://lindeloev.github.io/job/):

```{r, eval=F}
job::job({
  tag_download_wind(tag)
})
```

## Create graph

Similar to the example of the Swainson's Warbler in the basic tutorial, we first need to create the trellis graph:

```{r 11-graph-create, cache=TRUE, message=F, results='hide'}
graph <- graph_create(tag)
```

## Add wind to graph

We then compute the average windspeed experienced by the bird for each edge of the graph. This process can be quite long as we need to interpolate the position of the bird along its flight on a 4D grid (latitude-longitude-pressure level-time). 

We then compute the airspeed based on this windspeed and the known groundspeed. All of these are stored as [complex values](https://stat.ethz.ch/R-manual/R-devel/library/base/html/complex.html) with the real part representing the E-W component and the imaginary part corresponding to the N-S component.

```{r 11-add-wind, cache=TRUE, cache.lazy = FALSE, message=F, results='hide'}
graph <- graph_add_wind(graph, pressure = tag$pressure)
```

## Define movement model

While you can still define the movement model with a parametric function (i.e., gamma or logit), we will see how to use the mechanical power curve. The power curve expresses the energy required for a bird to fly at a certain airspeed based on aerodynamic theory. See more details in [section 2.2.5 of Nussbaumer et al. (2023b)](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.14082#mee314082-sec-0009-title).

First, we search for morphological information on the Great Reed Warbler using the [AVONET database](https://doi.org/10.6084/m9.figshare.16586228.v5).

```{r}
bird <- bird_create("Acrocephalus arundinaceus")
```

Using the bird created, we can set the movement model with

```{r}
graph <- graph_set_movement(graph,
  method = "power",
  bird = bird,
  power2prob = \(power) (1 / power)^3
)
plot_graph_movement(graph)
```

## Products

We can then compute the same three products as for the Swainson's Warbler:

```{r 11-products, cache = TRUE, message=F, results='hide'}
path_most_likely <- graph_most_likely(graph)
marginal <- graph_marginal(graph)
path_simulation <- graph_simulation(graph, nj = 10)
```

```{r}
plot(marginal, path = path_most_likely)
plot_path(path_simulation, plot_leaflet = F)
```

## Save

`graph` can become extremely big for such model and it might not be recommended to save it. Check its size with `format(object.size(graph), units = "MB")`.

```{r 11-save, warning=F}
save(
  tag,
  # graph,
  path_most_likely,
  path_simulation,
  marginal,
  file = "./data/interim/18LX.RData"
)
```
