# Introduction

[GeoPressureTemplate](https://github.com/Rafnuss/GeoPressureTemplate) is a [GitHub repository template](https://docs.github.com/articles/creating-a-repository-from-a-template/) which contains a standard folder structure and R scripts to store your data, analyse it, and produce trajectory outputs. Download it as a starting point to organise your data and start your analysis. It's an initial flexible structure that should be adapted to each individual project.

## 📁 Project structure

GeoPressureTemplate defines a standardized project folder structure based on a mix of [rrrpkg](https://github.com/ropensci/rrrpkg#getting-started-with-a-research-compendium), [rrtools](https://github.com/benmarwick/rrtools#4-rrtoolsuse_analysis) and [cookiecutter data science](http://drivendata.github.io/cookiecutter-data-science/#directory-structure).

```
GeoPressureTemplate/
├── DESCRIPTION                             # Project metadata and dependencies
├── README.md                               # Top-level description of content and guide to users
├── GeoPressureTemplate.Rproj               # R project file
├── LICENCE.md                             # Conditions of re/use the data and code
├── config.yml                              # YML file used to define the parameters used in the analysis
├── data/                                 
│   ├── raw_tag/                            # Raw geolocator data (do not modify!)
│   │   ├── 18LX/
│   │   │   ├── 18LX_20180725.acceleration
│   │   │   ├── 18LX_20180725.glf
│   │   │   └── 18LX_20180725.pressure 
│   │   └── CB619/ 
│   │       └── CB619.deg
│   ├── tag_label/                          # Trainset csv file generated with analysis/1-label.qmd
│   │   ├── 18LX-labeled.csv
│   │   ├── 18LX.csv 
│   │   ├── CB619-labeled.csv
│   │   └── CB619.csv            
│   ├── twilight_label/                     # Trainset csv file generated with analysis/2-twilight.qmd
│   │   ├── 18LX-labeled.csv
│   │   └── 18LX.csv
│   ├── wind/                               # ERA-5 wind data generated with analysis/3-wind.qmd
│   │   └── 18LX/
│   │       ├── 18LX_1.nc
│   │       └── ...
│   └── interim/                            # Intermediate data created with analysis/4-geopressuretemplate.R
│       └── 18LX.RData                      
├── analysis/                               # R scripts used to analyse your data.
│   ├── 1-label.qmd
│   ├── 2-twilight.qmd
│   ├── 3-label-advanced.qmd
│   ├── 10-geopressuretemplate.R
│   └── 20-datapackage.R
├── R/                                      # Custom R functions used in your analysis.
└── output/   
    ├── create_figures.R
    └── figures/
```

## 💡 Get started

### 🛠️ Create your project


**Option 1: with a Github repository (recommended)**

- Open [GeoPressureTemplate](https://github.com/Rafnuss/GeoPressureTemplate/) and click on the green button "Use this template" or [click here](https://github.com/Rafnuss/GeoPressureTemplate/generate) to create your repository directly.
- Choose a project name (e.g., `my_tracking_study_name`) specific to your research. Note that this becomes the name of your folder on your computer too.
- Clone the repository on your computer.
- Done! 🎉

**Option 2: using the GeoLocatoR package**

Alternatively, create your project folder with [`GeoLocatoR::create_geopressuretemplate()`](https://raphaelnussbaumer.com/GeoLocatoR/reference/create_geopressuretemplate.html).

```{r}
#| eval: false
library(GeoLocatoR)
create_geopressuretemplate(
  "location/to/your/project/folder/my_tracking_study_name"
)
```


**Option 3: Manual download**

- Click [here](https://github.com/Rafnuss/GeoPressureTemplate/archive/refs/heads/main.zip) to download the repo directly 
- Unzip and rename the folder with your project name.
- Done! 🎉

::: callout-important
Do not clone (or fork) the repository! [Read more about creating repositories from a template](https://docs.github.com/en/repositories/creating-and-managing-repositories/creating-a-repository-from-a-template).
:::


### 🏠 Make yourself at home

1. Rename `GeoPressureTemplate.Rproj` to your project name (i.e., `my_tracking_study_name.Rproj`). You can now open the project in Positron/RStudio.
2. Edit the `DESCRIPTION` file following the [R package book](https://r-pkgs.org/description.html)
3. Replace the content of `data/` with your tag data.
4. Install the dependencies needed with

``` r
devtools::install()
```

## 💫 How to master `config.yml` ?

Before jumping into the analysis, we need to introduce `config.yml`. It's a [YAML file](https://en.wikipedia.org/wiki/YAML) which defines the parameters used in the analysis. Separating these parameters from the main code follows [best practices](https://en.wikipedia.org/wiki/Separation_of_concerns) in terms of reproducibility, readability, and sharability. We use the [config](https://rstudio.github.io/config/) R package to retrieve all the parameter values used in the code.

::: callout-note
An important benefit of `config` is its ability to manage hierarchical configurations: you can organize settings into different groups and override them as needed. This is particularly useful in our context, where we can configure parameters for all tags and/or specific tags.
:::

``` yml
default:
  tag_create:
    extent: [-16, 23, 0, 50] # [W, E, S, N]
    scale: 1

18LX:
  tag_set_map:
    scale: 5
```

``` r
config::get("tag_create", "18LX")$extent
# [-16, 23, 0, 50] # Returns the default value if not set in 18LX
config::get("tag_set_map", "18LX")$scale
# 5 # Return the value in 18LX if presents in both
```

## 📈 Analyse the data

Now that you're set up, it's time to get serious 😬.

### Step 1: Preparation

Here we make sure everything is ready to run the model. This mainly involves defining the parameters in `config.yml` by running the following three scripts:

In this part, we will be presented step by step how to analyse your data:

1. Run [`analysis/1-label.qmd`](/geopressuretemplate-label.html).
2. Run [`analysis/2-twilight.qmd`](/geopressuretemplate-twilight.html), if you have light data.

You can visualize each of the scripts in a nice-to-read HTML format, using the example of 16LX, in the next sections of this manual.

::: callout-note
When running the scripts, please keep in mind:

- Nothing is saved at the end of these scripts (and that's how it's supposed to be!). Only label files and `config.yml` should be edited.
- The scripts should be run for each tag separately.
- We use [Quarto](https://quarto.org/) to make it easy to run chunks based on your needs (e.g., re-run a small chunk after making a change). The scripts are *not* meant to be run with `Run all`.
- These scripts can (and should) be edited based on your prefered workflow.
:::

::: callout-tip
For regular users, we provide a shorter version of these preparation steps in `analysis/3-label-advanced.qmd`.
:::

### Step 2: Run GeoPressureR workflow

::: callout-caution
The core benefit of using GeoPressureTemplate is coming now! You can finally gain back all the time you've invested in setting up this project folder! Keep reading carefully!
:::

The `geopressuretemplate()` function allows you to run the entire GeoPressureR workflow with a single function. By using this workflow, you make sure that your work is reproducible.

```{r 09-workflow-run}
#| eval: false
geopressuretemplate("18LX")
```

This creates `data/interim/18LX.RData` with all the trajectory outputs. You'll learn more in the [Workflow chapter](/geopressuretemplate-workflow.html). In GeoPressureTemplate, this is performed in the file `analysis/4-geopressuretemplate.R`.

### Step 3: Your own analysis

You've reached the final steps of GeoPressureTemplate, but this is only the beginning of your research journey. You now have the bird trajectories, but the real magic happens when you apply your creativity and insights to analyse these! You can continue using GeoPressureTemplate, adding other sources of data and analysis to answer your research question.  


## 🗞️ Publication

For peer-review publication, it is essential that the data and code are accessible to reviewers. Because inaccurate labelling can lead to a wrong trajectory, we highly encourage you to publish your data and code.

### GeoPressureTemplate Code

Simply publish your GeoPressureTemplate Github repository on Zenodo following [this guide](https://docs.github.com/en/repositories/archiving-a-github-repository/referencing-and-citing-content). Publishing on Zenodo will generate a DOI for your code.

### Geolocator Data Package

While publishing your GeoPressureTemplate on Zenodo is a great step to allow others to see (and, if needed, reproduce) your work, it can be a challenge to re-use your track data for other purposes. To facilitate this, we created [GeoLocator Data Package](https://raphaelnussbaumer.com/GeoLocator-DP/), a data exchange format for geolocator data, providing all GeoPressureR outputs in csv files. The aim of GeoLocator DP is re-usability rather than reproduction. 

This step is performed with the script `analysis/5-datapackage.R`. To learn more about publishing a GeoLocator Data Package, go to [GeoLocatoR DP](geolocator-intro.html).


## 🔗 Advanced options

- Generate a citation file with [`usethis::use_citation()`](https://usethis.r-lib.org/reference/use_citation.html) or [`cffr`](https://github.com/ropensci/cffr).
- Use [`renv`](https://rstudio.github.io/renv) to make your work reproducible.
