# GeoPressureViz

```{r 07-setup}
#| message: false
library(GeoPressureR)
```

GeoPressureR includes the <a href="https://shiny.rstudio.com/" target="_blank">shiny app</a> `GeoPressureViz` which helps you visualize the overall trajectory of the bird as well as each step-by-step move. Note that **this is a visualization tool only** and as such, it cannot be used to make changes to your data. 

It is an essential tool to label your tag and provides a convenient platform to discuss the overall trajectory with collaborators.

{{< video https://www.youtube.com/embed/x1WDPIiFB5g >}}

## Run GeoPressureViz

Start the shiny app with

```{r 07-geopressureviz-run}
#| eval: false
geopressureviz(tag)
# or if interim file is present
geopressureviz("18LX")
```


## Navigate GeoPressureViz


1.  The app has three panels:

    1.  **Map**: this displays the trajectory of the bird. The size of the circles indicates the duration of the stopover (hover over them to get more info).

    2.  **Bottom panel**: this shows the `pressurepath` of the trajectory which must be computed beforehand. The grey line is the raw data (`pressurepath$pressure_tag`), the black dots are pressure points labelled as outliers, and colored lines are the normalized pressure at the best match location (i.e., `pressurepath$surface_pressure_norm`). The color matches the circles on the map.

    3.  **Side panel**: this provides key information and tools to navigate the app.

2.  The "Minimum duration" input allows you to filter out short stationary periods on the map, similarly to `include_min_duration` in `tag_set_map()`.

3.  Toggle the "Full Track" button to move to the stationary period view.

4.  Navigate from one stationary period to the next using the previous/next button or dropdown list. The pressure timeseries will update to zoom exactly on this stationary period. The color of the timeseries informs you about the current stationary period shown on the map, the two others being the previous and next one.

5.  The circles indicates the distance to and from the previous and next stationary period location based on the speed slider on the side panel (default is 40km/h).

6.  Depending on which maps are available, you can choose to display the probability map of pressure `tag$map_pressure`, light `tag$map_light`, their product, or the marginal.

7.  Because some locations of the most likely path can be completely off, it might be helpful to temporarily change their location to check the coherence of distance with the next/previous stationary period. Click on "Start Editing" to do so. Every time you click on the map, it updates the position. Once you're happy with the position, you can click on the button again to disable the editing mode.

8.  Finally, you can also check the pressure timeseries with the new positions using the button "Query pressure". This runs `geopressure_timeseries()` in the background, so it is slow (especially for long stationary periods).

Note that the path of the shiny app can be retrieved from the global environment using `geopressure_path`.
