# Trajectory with wind

```{r, 11-setup-1}
#| message: false
#| include: false
library(GeoPressureR)
library(ggplot2)
```

```{r 11-setup-2}
#| cache: true
#| include: false
tag <- tag_create("18LX",
  crop_start = "2017-06-20",
  crop_end = "2018-05-02",
  quiet = TRUE
) |>
  tag_label(quiet = TRUE) |>
  tag_set_map(
    extent = c(-16, 23, 0, 50),
    scale = 4,
    known = data.frame(
      stap_id = 1,
      known_lat = 48.9,
      known_lon = 17.05
    )
  ) |>
  geopressure_map(quiet = TRUE) |>
  twilight_create() |>
  twilight_label_read() |>
  geolight_map()
```

In this second chapter of the advanced tutorial, we explore how to model the trajectory of the Great Reed Warbler using wind data.

Wind can significantly influence a bird's movement, explaining up to 50% of the displacement! Accounting for wind allows us to estimate the airspeed of each transition rather than groundspeed. As such, the movement model can be defined as the probability of a bird's airspeed, which is much more constrained and precise. This approach is presented in detail in <a href="https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.14082#mee314082-sec-0008-title" target="_blank">section 2.2.4</a> of @Nussbaumer2023b.

## Download wind data

Wind data is available at high resolution (1hr, 0.25°, 37 pressure level) on ERA5 hourly data on pressure levels [@copernicusclimatechangeservice2018]. This data is easily accessible through the <a href="https://bluegreen-labs.github.io/ecmwfr" target="_blank">`ecmwfr`</a> package.

::: callout-caution

Learn how to set-up your ECMWF account to be able to download ERA5 data on the [GeoPressureTemplate/Download Wind page](geopressuretemplate-wind.html#set-copernicus-credentials).

:::

As the flights tend to be of short duration, we suggest downloading a file for each flight. This can be done automatically with `tag_download_wind()`, which uses <a href="https://bluegreen-labs.github.io/ecmwfr/articles/advanced_vignette.html#batch-parallel-requests" target="_blank">`wf_request_batch()`</a> to make all the requests in parallel.

```{r} 
#| eval: false
tag_download_wind(
  tag,
  variable = c("u_component_of_wind", "v_component_of_wind", "temperature")
)
```

You can monitor the requests at [https://cds.climate.copernicus.eu/requests]. The files are downloaded in `data/wind/`

::: callout-note

In addition to the two required wind variables, we also downloaded the temperature data during the flights. This will later allow us to retrieve temperature data at the exact location of the bird during the flight. See `tag_download_wind()` documentation for more information on the available variables. 

:::


::: callout-tip

In case you have a lot of tracks for which you need to download wind data and don't want to block your console, you might consider using an <a href="https://solutions.rstudio.com/r/jobs/" target="_blank">RStudio background job</a>, which can be easily called with the <a href="https://lindeloev.github.io/job/" target="_blank">job package</a>:

```{r} 
#| eval: false
job::job({
  tag_download_wind(tag)
})
```

:::

## Create graph

Similar to the example of the Swainson's Warbler in the basic tutorial, we first need to create the trellis graph:

```{r 11-graph-create}
#| cache: true
graph <- graph_create(tag, quiet = TRUE)
```

## Add wind to graph

We then compute the average windspeed experienced by the bird for each edge of the graph. This process can be quite long as we need to interpolate the position of the bird along its flight on a 4D grid (latitude-longitude-pressure level-time).

We then compute the airspeed based on this windspeed and the known groundspeed. All of these are stored as <a href="https://stat.ethz.ch/R-manual/R-devel/library/base/html/complex.html" target="_blank">complex values</a> with the real part representing the E-W component and the imaginary part corresponding to the N-S component.

```{r 11-add-wind}
#| cache: true
graph <- graph_add_wind(graph, pressure = tag$pressure, quiet = TRUE)
```

## Define movement model

While you can still define the movement model with a parametric function (i.e., gamma or logit), we find it more intuitive to use the mechanical power curve. The power curve expresses the energy required for a bird to fly at a certain airspeed based on aerodynamic theory. See more details in <a href="https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.14082#mee314082-sec-0009-title" target="_blank">section 2.2.5</a> of @Nussbaumer2023b.

First, we search for morphological information on the Great Reed Warbler using the <a href="https://doi.org/10.6084/m9.figshare.16586228.v5" target="_blank">AVONET database</a> [@tobias2022].

```{r}
bird <- bird_create("Acrocephalus arundinaceus")
```

Using the bird created, we can set the movement model by converting the airspeed to power, and power to a probability. This second step is still a parametric equation, which can be manually defined with `power2prob`.

```{r}
graph <- graph_set_movement(graph,
  method = "power",
  bird = bird,
  power2prob = \(power) (1 / power)^3,
  low_speed_fix = 15
)
plot_graph_movement(graph)
```

## Products

We can then compute the same three products as for the Swainson's Warbler:

```{r 11-products}
#| cache: true
path_most_likely <- graph_most_likely(graph, quiet = TRUE)
marginal <- graph_marginal(graph, quiet = TRUE)
path_simulation <- graph_simulation(graph, nj = 10, quiet = TRUE)
```

```{r}
plot(marginal, path = path_most_likely)
```

```{r}
plot_path(path_simulation, plot_leaflet = FALSE)
```

### Extract flight information

The `path`variable contains all the information at the scale of the stationary period. However, to get flight information, you need to extract variable of the edge of the graph. `path2edge()` is the function for that!

```{r}
edge_simulation <- path2edge(path_simulation, graph)
edge_most_likely <- path2edge(path_most_likely, graph)
knitr::kable(head(edge_most_likely, 3), digits = 1)
```

```{r}
#| echo: false
edge_simulation$name <- glue::glue("{edge_simulation$stap_s}->{edge_simulation$stap_t}")
plot1 <- ggplot(edge_simulation, aes(reorder(name, stap_s), abs(gs))) +
  theme_bw() +
  ylab("gs") +
  geom_boxplot()
plot2 <- ggplot(edge_simulation, aes(reorder(name, stap_s), abs(ws))) +
  theme_bw() +
  ylab("ws") +
  geom_boxplot()
plot3 <- ggplot(edge_simulation, aes(reorder(name, stap_s), abs(gs - ws))) +
  theme_bw() +
  ylab("as") +
  geom_boxplot()
plot4 <- ggplot(edge_simulation, aes(reorder(name, stap_s), duration)) +
  theme_bw() +
  ylab("duration") +
  geom_point()

plotly::subplot(
  plotly::ggplotly(plot1),
  plotly::ggplotly(plot2),
  plotly::ggplotly(plot3),
  plotly::ggplotly(plot4),
  nrows = 4, titleY = TRUE
)
```

It's a good idea to check the distribution of ground speed (gs), winspeed (ws) and airspeed (as) and check for any outliers which might come from error in the labelling. Here you can see the high groundspeed (\>100km/h between stap `24` and `25`) which is nicely explained by wind, as the corresponding airspeed is perfectly normal (\~50km/h)

We can (and should) also check that our movement model is coherent with the distribution of flight speed assumed in the movement model:

```{r}
plot_graph_movement(graph) +
  geom_histogram(
    data = data.frame(as = abs(edge_simulation$gs - edge_simulation$ws)),
    aes(x = as, y = after_stat(count) / sum(after_stat(count))),
    color = "black", fill = NA, bins = 100
  )
```

If you find anomalous flight speed, it might be worth checking if this/these flight(s) have been correctly labelled.

## Save

`graph` can become extremely big for such models and it might not be recommended to save it. Check its size with `format(object.size(graph), units = "MB")`.

```{r 11-save}
#| warning: false
save(
  tag,
  graph,
  path_most_likely,
  path_simulation,
  marginal,
  edge_simulation,
  edge_most_likely,
  file = "./data/interim/18LX.RData"
)
```
