# Trajectory

```{r 03-setup}
#| message: false
#| code-fold: true
library(GeoPressureR)

tag <- tag_create(id = "CB619") |>
  tag_label() |>
  tag_set_map(
    extent = c(-100, -68, 0, 40), # coordinates of the map to request (W, E, S, N)
    scale = 2, # request on a 1/2=0.5° grid, coarse, but fast
    known = data.frame(
      stap_id = 1,
      known_lat = 37.286812,
      known_lon = -82.304972
    )
  ) |>
  geopressure_map(
    max_sample = 100,
    margin = 20,
    thr_mask = 0.95,
    sd = 0.5,
    log_linear_pooling_weight = \(n) 4 * log(n) / n,
    keep_mse = TRUE,
    quiet = TRUE
  )
```

In this last chapter of the basic tutorial, we model the full trajectory of the Swainson's Warbler by combining the observation model of pressure with a movement model of flight accounting for likely speed to constrain possible distances between stationary periods. To do so, we use a highly optimized <a href="https://en.wikipedia.org/wiki/Hidden_Markov_model" target="_blank">Hidden Markov model</a> presented in @Nussbaumer2023b.

This involves three main steps:

1.  **Create the graph** representing the possible trajectory of the bird.
2.  Set the **movement model** which defines the likelihood of each possible flight encoded in the graph.
3.  **Compute three products** of the model: (1) the most likely trajectory, (2) the probability maps at each stationary period (i.e., marginals), and (3) simulated paths sampled according to their likelihood.

## Background on movement modeling

If you haven't heard of <a href="https://en.wikipedia.org/wiki/Hidden_Markov_model" target="_blank">Hidden Markov model</a>, <a href="https://doi.org/10.1016/j.tree.2007.10.009" target="_blank">state-space models</a>, or <a href="https://en.wikipedia.org/wiki/Trellis_(graph)" target="_blank">trellis graphs</a>, I recommend that you start by reading the summary below, or - even better - read @Patterson2008 and @Nussbaumer2023b if you're feeling brave!

<details>

<summary>Click here to read a short summary of movement model</summary>

::: callout-note
We start by making the fundamental assumption that the overall trajectory can be separated into **discrete** positions which, together, create a trajectory.

This discretization allows us to use a <a href="https://en.wikipedia.org/wiki/Hidden_Markov_model" target="_blank">Hidden Markov model</a>, which conveniently distinguishes the **unknown positions** of a bird $X_t$ from some **known observations** $Y_t$ which can be related to these positions (e.g., pressure or light measurement). Defining an HMM requires that we set:

1.  An **observation model** $P(Y_t | X_t)$, which defines the likelihood of observing a measurement at a given position. This is exactly what our pressure likelihood map computed in the previous chapter does.
2.  A **movement model** $P(X_t+1|X_t)$, defining the probability that a bird was able to fly a specific distance between two consecutive stationary periods.
3.  An **initial state** $P(X_0)$, technically required but this does not constitute a strong constraint in our case.

Let's say that our goal is to produce the probability map for each stationary period accounting for all pressure measurements while maintaining likely flights. This corresponds to the **marginal** probability map $P(X_t=x_t|Y_0=y_0,...Y_t=y_t)$. If you are not familiar with the lower and capital case, $X_t$ represents the random variable (i.e., the position of the bird at a stationary period $t$), while $x_t$ is a specific value that $X$ can take, in our case, the specific coordinates of pixels on the map.

Because the number of possible positions is limited (i.e., number of pixels on the map), we can solve the HMM exactly. This is in contrast to Markov Chain Monte Carlo (MCMC) algorithm, which does not assume any discretization of space but relies on an iterative approach to solve the model.

A graph is a convenient mathematical representation of a model which has states (nodes) and relations between these states (edges). Graphs are a general and fundamental mathematical tool, but here we use a specific graph structure corresponding to the HMM presented above: two states $X$ and $Y$ for each time step $t$ (nodes), and an observation and movement model (relating $X_t$ to $X_t+1$ and $Y_t$ to $X_t$ respectively). All other relationships are assumed to be independent.

Typically in animal movement, the graph is only used to represent the random variable of the HMM, but in our case, we also use the concept of graph to store all discrete positions (i.e., pixels) of the bird at all timesteps, referred to as a <a href="https://en.wikipedia.org/wiki/Trellis_(graph)" target="_blank">trellis graphs</a>.

You can think of the nodes as a full 3D grid of lat x lon x stationary period and the edges as all possible pairs of nodes between two consecutive stationary periods. As you can imagine, this results in a very high number of edges! The novelty of @Nussbaumer2023b is to construct this graph using only the useful nodes and edges, thereby dramatically reducing the computational resources required.
:::

</details>

## Create the graph

We create the trellis graph with the function `graph_create()`, which performs the following steps:

1.  Keep only the nodes of all locations within a threshold percentile of the likelihood map `thr_likelihood`. For instance, a value of `0.90` means that We only keep the pixels for which their likelihood sums represent 90% of the total probability space.
2.  Keep only the nodes which are within reach of at least one node of the previous and one of the next stationary period. This is done by assuming a maximal possible groundspeed `thr_gs`. This key step is efficiently computed with the image binary distance function `EBImage::distmap()`.
3.  Build the edges based on the filtered nodes and the groundspeed threshold `thr_gs`.
4.  Prune the graph based on the flow graph constraint, that is, that all nodes must be connected to the equipment node(s) and retrieval node(s), to avoid dead branches (where some nodes are connected in a single direction).

These steps are explained and illustrated in more detail in [section 2.2](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.14082#mee314082-sec-0004-title) of @Nussbaumer2023b

```{r 03-graph-create}
#| cache: true
graph <- graph_create(
  tag,
  thr_likelihood = 0.90,
  thr_gs = 100, # km/h
  quiet = TRUE
)
```

## Set movement model

We then define the movement model by setting a parametric function converting groundspeed to probability.

In this basic tutorial, we use the gamma distribution to model the probability of groundspeed.

```{r 03-graph-set-movement}
graph <- graph_set_movement(
  graph,
  method = "gamma",
  shape = 7,
  scale = 7,
  low_speed_fix = 15
)
```

Because birds may fly back and forth over small distances, we give all groundspeeds under 15km/h the same probability as for 15 km/h with the parameter `low_speed_fix`.

We can visualize the parametric function created with `plot_graph_movement()`

```{r 03-graph-movement-plot}
plot_graph_movement(graph)
```

## Products

With the graph built, we can now generate the three main products of GeoPressureR, presented in more detail in [section 2.3](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.14082#mee314082-sec-0010-title) of @Nussbaumer2023b.

### Product 1: Most likely path

The graph can directly and efficiently compute the most likely path, that is, the set of positions at each stationary period which maximizes the joint probability of the trajectory. It finds the optimal compromise between the most likely position according to the pressure (and light) likelihood map and the realistic flight distance defined by the movement model.

```{r 03-path-most-likely-create}
path_most_likely <- graph_most_likely(graph, quiet = TRUE)
```

In GeoPressureR, a `path` is a standard data.frame created and used [in several other functions](https://raphaelnussbaumer.com/GeoPressureR/reference/index.html#path).

```{r 03-path-most-likely-summary}
knitr::kable(head(path_most_likely, 3))
plot_path(path_most_likely)
```

### Product 2: Marginal probability map

Estimating the uncertainty of the bird's position for each stationary period is generally the most sought-after output of tracking studies. While the pressure likelihood map provides some information for each stationary period, we can now produce the probability map of positions accounting for all observations and transitions. In the HMM, these maps correspond to the marginal distribution of a Markov process.

```{r 03-map-marginal-create}
marginal <- graph_marginal(graph, quiet = TRUE)
```

You can visualize the marginal map along with the most likely path:

```{r 03-map-marginal-plot}
plot(marginal, path = path_most_likely)
```

### Product 3: Simulated paths

When you want to perform a post-hoc computation on the trajectory (e.g., mean flight distance), it is convenient to simulate multiple paths, compute the metric on each of them (e.g., flight distance) and compute the summary statistics afterwards (e.g., mean).

Note that we are using the forward filtering backward sampling algorithm, allowing to sample trajectories independently and without error.

```{r 03-path-simulation-create}
path_simulation <- graph_simulation(
  graph,
  nj = 10, # Number of simulation
  quiet = TRUE
)
```

```{r 03-path-simulation-plot}
plot_path(path_simulation, plot_leaflet = FALSE)
```

## Save

You might want to save some of the key objects such as `tag` or `marginal` in an `R` format. Following the default structure described in [GeoPressureTemplate](https://github.com/Rafnuss/GeoPressureTemplate#project-structure-file_folder), you should save them in the `./data/interim` folder.

Note that `graph` can be quite large if the number of stationary periods and grid resolution is high.

```{r 03-save-interim}
#| warning: false
save(
  tag,
  graph, # format(object.size(graph), units = "MB")
  path_most_likely,
  path_simulation,
  marginal,
  file = "./data/interim/CB619.RData"
)
```

When saving `tag` and `graph`, you also archive the functions parameters used, which can be retrieved with `tag$param` and `graph$param`.
